
# About the project

*Github repository : https://github.com/kbatpurev/IODS-project.git This course seems like a good fit for me. Even though i am familiar with most of the topics listed, i always find gaps in my knowledge and technique as i am self-taught R student. The bit that i am looking forward to the Git and Rmarkdown part a lot. I am a complete novice on both, but i am planning to get into it with sleeves up.I learnt about the course through Sisu. RHDS is a pretty good option for a crash course i feel. I liked the exercises. Favourite chapter was 4, data visualisation.*

# IODS 23

## Main learning outcomes
Git and Rmarkdown. And filling the gaps in my knowledge

## Favourite chapter from RHDS book
Chapter 4 Data Vis

## Source of info about course
Sisu


```{r}

```

## Getting familiar with gapminder data

```{r}
date()
library(tidyverse)
library(gapminder)

head(gapminder)
gapminderMongolia=gapminder %>% filter(country=="Mongolia")
#Only 4 fold increase in Mongolia's GPD over 55 years between 1952 and 2007. 
#Lets see what Australia looks like
gapminderAustralia=gapminder %>% filter(country=="Australia")
#Only just over 3 folds! Less than Mongolia=Very interesting. 
gapminder1997=gapminder%>%filter(year==1997)
gapminder1997%>%ggplot(aes(x=gdpPercap,y=lifeExp))+
  geom_point()

t.test(lifeExp~gdpPercap>20000,data=gapminder1997)
#Life expectancy in countries with gpd above 20000 (i am guessing USD) is higher than poorer countries. Expected
```

